Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2023-12-13 ms.qmd
Updated MS. Intro roughly finished

---
title: "Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

<p style="text-align: center;">Daniel W.A. Noble$^1$, Fonti Kar$^2$, Frank Seebacher$^3$, Alex Bush$^4$, & Shinichi Nakagawa$^2$</p>

#### Affliations:
$^1$ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
$^2$ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
$^3$ SOLES, University of Sydney, Sydney, NSW, Australia &NewLine;  
$^4$ Department of Biology, Lancaster University, Liverpool, UK

@@

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation may also have important ecological consequences by promotong population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011]. The implications of changes in variance could therefore have wide-reaching consequences for understanding the capacity for populations to persist in and adapt to novel environments [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., see @Seebacher2015-wi], it is unclear if the magnitude of physiological adjustment is sufficient. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits increases in temperature may have a stronger impact on variation in physiological rates within populations compared to aqautic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the impacts ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that the opportunity for selection is expected to decrease for ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide a more complete picture of the impacts that climate warming will have on ectotherm populations, with important ecological and evolutionary implications that require further attention.

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key]. Higher spatial heterogenetiy in terrestrial ectotherms suggests that plasticity should be more likely evolve, however, we did not find evidence to this effect; possibly because of the lower predictability of thermal environments over time making cues unreliable [@hendry2016key]. Empirical evidence properly testing these predictions, however, is scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theoretical explanations may not be general but rather specific to the population, context and trait in question; a point emphaised by others.

### ***Ecological consequences of reduced variation in physiological Rates***
Variance in physiological rates decreases substantially at high temperatures with effects being especially pronouced in terrestrial ecosystems and for key traits such as metabolism and performance. Changes in variability in physiological rates, particularly energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems that can have important consequences for population dynamics [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increasdd growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. For examaple, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of authorities . To date, however, an appreciation for the wide-reaching impacts changes in variance may have on populations and communities remains limited. Indeed, the ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. Such insight will require the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models.

>>>>

---
title: "Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/ecology-letters.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

<p style="text-align: center;">Daniel W.A. Noble^1^, Fonti Kar^2^, Frank Seebacher^3^, Alex Bush^4^, & Shinichi Nakagawa^2^</p>

#### Affliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ SOLES, University of Sydney, Sydney, NSW, Australia &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK

@@

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Dewitt1998; @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020; @Dewitt1998]. Such changes in physiological trait variation may also have important ecological consequences by promotong population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011]. The implications of changes in variance could therefore have wide-reaching consequences for understanding the capacity for populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn; @hoffmann2013upper]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits increases in temperature may have a stronger impact on variation in physiological rates within populations compared to aqautic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are as of yet poorly measured and modelled.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that variance in physiological rates is expected to decrease for ectotherms as temperatures increase, but particularly so for terrestrial ectotherms (~28%). Decreases in variance of physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide a more complete picture of the impacts that climate warming will have on ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention.

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key]. Higher spatial and temporal heterogenetiy in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely evolve, however, we did not find evidence to this effect; possibly because of the lower predictability of thermal environments over time making cues unreliable [@hendry2016key]. Empirical evidence properly testing these predictions is also scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theoretical explanations may not be general but rather specific to the population, context and trait in question; a point emphaised by others.

### ***Ecological consequences of reduced variation in physiological Rates***
Variance in physiological rates decreases substantially at high temperatures with effects being especially pronouced in terrestrial ecosystems and for key traits such as metabolism and performance. Changes in variability in physiological rates, particularly energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems that can have important consequences for population dynamics [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. For examaple, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of authorities . To date, however, an appreciation for the wide-reaching impacts changes in variance may have on populations and communities remains limited. Indeed, the ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. Such insight will require the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models.
